\textbf{On the Common Logical Structure of Classical and Quantum Mechanics}

### 2021-05-14 Articles/2020-QuantumLogic/Source/quantumlogic.tex
QLogic: minor comments for Andrea

From a mathematical perspective, the distributive lattice will be provided in both classical and quantum case by a Borel algebra, which is the smallest $\sigma$-algebra on a set $X$ equipped with a topological structure that contains the topology; the non-distributive lattice will be provided by the set of subgroups ordered by inclusion. We will claim that the non-distributive lattice is a strict subset of the distributive one; this result entails that everything that can be described in quantum logic can also be described in classical logic. Moreover, the distributive lattice will include propositions, like sentences concerning the expectation value of observables that are of high scientific interest, since in an experimental context by ``measurement'' one typically refers to the statistical collection of many ``single takes''. On the contrary, it is well-known that standard quantum logic deals only with single-take measurements, remaining silent w.r.t.\ propositions involving statistical statements. Hence, measurements of cross sections, the most typical type of measurement in the context of particle physics, are not part of the quantum lattice, and neither are measurements of the probabilities themselves. Then, to show that one may apply a distributive lattice to quantum propositions can be seen as an important result in order to extend the set of meaningful logical propositions of quantum mechanics. To provide an extended quantum logic lies beyond the introductory scope of the present paper and will be the subject for future research; here we limit ourselves to show that such an extension is possible.

@@

As anticipated above, in this section our arguments are fundamentally based on the notion of $\sigma$-algebra, which is a significant structure in the context of foundational research, given that physically relevant mathematical branches like measure theory (cf.\ \cite{Cohn:2013}) and probability theory (cf.\ \cite{Grimmett:2001}) are built on top of it. Let us then review some of its essential features.

@@

In the second place, it is crucial to underline that every topological space---as for instance the complex Hilbert space of quantum mechanics and the real Hilbert space of classical probability distributions considered in this essay---comes already equipped with a $\sigma$-algebra called Borel algebra.\ A topology, in fact, is also a collection of sets---called \emph{open sets}---and the Borel algebra is defined to be the smallest $\sigma$-algebra that contains all the open sets.\footnote{Here the topology is important for us only as a mean to construct the Borel algebra.} Complete normed spaces, such as the mentioned Hilbert spaces, have a canonical way to construct a topology and, therefore, a $\sigma$-algebra. Let us see how this procedure works in the easier case of a standard three-dimensional Euclidean space. Given two points $x$ and $y$, we can write the square of distance $\epsilon^2 = |y - x|^2$ as the square of the norm of the vector between the two points. The statement ``the position is within $\epsilon$ meters of $x$'' corresponds to the set $U = \{y \; | \; |y - x|^2 < \epsilon^2 \}$ of all the points whose distance from $x$ is less than $\epsilon$. These are exactly the open sets that generate the topology, and the Borel algebra will be the smallest $\sigma$-algebra that contains these sets. This works analogously for all normed spaces. A Hilbert space, in particular, comes with an inner product which also defines a norm. We can define a distance $\epsilon^2 = |\psi - \phi|^2=\langle \psi - \phi , \psi - \phi \rangle$ between two states, and the statement ``the state of the system is within $\epsilon$ of $\psi$'' corresponds to the set $U = \{\phi \; | \; |\psi - \phi|^2 < \epsilon^2\}$. As we have already stated, the space of classical probability distributions is also a Hilbert space where each point corresponds to the square root of the density.\footnote{The choice of the square root of the density over the density itself is merely a technical choice.} In this case we have a distance $\epsilon^2 = |\sqrt{\rho_1} - \sqrt{\rho_2}|^2=\langle \sqrt{\rho_1} - \sqrt{\rho_2} , \sqrt{\rho_1} - \sqrt{\rho_2} \rangle$ and we can proceed in the same exact way. In all normed spaces, the topology is constructed from sets of the form ``the object is within $\epsilon$ of $x$''.

@@

Proof. Let $p \in L(H)$ be a closed sub-space of $H$. Then $p$ is a closed set, the complement of an open set, and therefore a Borel set: $p \in \Sigma$. Let $v \in H$ be a vector in the Hilbert space and consider the singleton $U = \{ v \}$. This is not a closed sub-space and therefore $U \notin L(H)$. \textcolor{red}{As the topology of a Hilbert space is Hausdorff, all singletons are closed and therefore $U \in \Sigma$. This means $L(H) \subset \Sigma$. }This concludes the proof.

@@

Now that we have briefly seen the definitions and basic constructions, we can gain more insights into these structures and what they represent. First, let us understand why the lattice of subspaces (i.e. quantum logic) fails to be a Boolean algebra. Stone's representation theorem tells us that any Boolean algebra can be expressed as an algebra of sets using the standard set operations (i.e. intersection, union and complement). For the lattice of subspaces, note that the disjunction $p \vee q$ does not correspond to the set union of the subspaces, but the span. In fact, the union of subspaces is not part of the lattice.  This is why the lattice, even if it is a complemented $0-1$ lattice, fails to be a Boolean algebra: it does not use the standard set operations.

@@

We want to be clear that we are not claiming that the lattice of subspaces of quantum logic is not of interest, or that it is of equal interest in both classical and quantum mechanics. Here we have pointed it out that the argument that quantum mechanics \emph{requires} necessarily a radical departure from classical logic is untenable, given that both theories are built on similar structures. We conclude that such claims should be substantially weakened, in the light of the arguments that we proposed in this section.

>>>>

From a mathematical perspective, the distributive lattice will be provided in both classical and quantum case by a Borel algebra, which is the smallest $\sigma$-algebra on a set $X$ equipped with a topological structure that contains the topology; the non-distributive lattice will be provided by the set of \textbf{GC: closed} subgroups ordered by inclusion. We will claim that the non-distributive lattice is a strict subset of the distributive one; this result entails that everything that can be described in quantum logic can also be described in classical logic. Moreover, the distributive lattice will include propositions, like sentences concerning the expectation value of observables that are of high scientific interest, since in an experimental context by ``measurement'' one typically refers to the statistical collection of many ``single takes''. On the contrary, it is well-known that standard quantum logic deals only with single-take measurements, remaining silent w.r.t.\ propositions involving statistical statements. Hence, measurements of cross sections, the most typical type of measurement in the context of particle physics, are not part of the quantum lattice, and neither are measurements of the probabilities themselves. \textbf{GC: La costruzione non \'e chiara... forse: Therefore, showing} that one may apply a distributive lattice to quantum propositions can be seen as an important result in order to extend the set of meaningful logical propositions of quantum mechanics. \textbf{GC: Providing} an extended quantum logic lies beyond the introductory scope of the present paper and will be the subject for future research; here we limit ourselves to show that such an extension is possible. \textbf{GC: Non ci \'e chiaro cosa voglia dire... ma forse \'e solo una frase tattica per dire `non aspettative che risolviamo tutto'?}

@@

As anticipated above, in this section our arguments are fundamentally based on the notion of $\sigma$-algebra, which is a significant structure \textbf{GC: from a foundational perspective (non era chiaro cosa foundational research sia... ma se \'e un termine tecnico, ricambia pure)}, given that physically relevant mathematical branches like measure theory (cf.\ \cite{Cohn:2013}) and probability theory (cf.\ \cite{Grimmett:2001}) are built on top of it. Let us then review some of its essential features.

@@

In the second place, it is crucial to underline that every topological space---as for instance the complex Hilbert space of quantum mechanics and the real Hilbert space of classical probability distributions considered in this essay---comes already equipped with a $\sigma$-algebra called Borel algebra.\ A topology, in fact, is also a collection of sets---called \emph{open sets}---and the Borel algebra is defined to be the smallest $\sigma$-algebra that contains all the open sets.\footnote{Here the topology is important for us only as a mean to construct the Borel algebra.} Complete normed spaces, such as the mentioned Hilbert spaces, have a canonical way to construct a topology and, therefore, a $\sigma$-algebra. Let us see how this procedure works in the easier case of a standard three-dimensional Euclidean space. Given two points $x$ and $y$, we can write the square of distance $\epsilon^2 = |y - x|^2$ as the square of the norm of the vector between the two points. The statement ``the position is within $\epsilon$ meters of $x$'' corresponds to the set $U = \{y \; | \; |y - x|^2 < \epsilon^2 \}$ of all the points whose distance from $x$ is less than $\epsilon$. These are exactly the open sets that generate the topology, and the Borel algebra will be the smallest $\sigma$-algebra that contains these sets. This works analogously for all normed spaces. A Hilbert space, in particular, comes with an inner product which also defines a norm. We can define a distance $\epsilon^2 = |\psi - \phi|^2=\langle \psi - \phi , \psi - \phi \rangle$ between two states, and the statement ``the state of the system is within $\epsilon$ of $\psi$'' corresponds to the set $U = \{\phi \; | \; |\psi - \phi|^2 < \epsilon^2\}$. As we have already stated, the space of classical probability distributions is also a Hilbert space where each point corresponds to the square root of the density.\footnote{The choice of the square root of the density over the density itself is merely a technical choice.} In this case we have a distance $\epsilon^2 = |\sqrt{\rho_1} - \sqrt{\rho_2}|^2=\langle \sqrt{\rho_1} - \sqrt{\rho_2} , \sqrt{\rho_1} - \sqrt{\rho_2} \rangle$ and we can proceed in the same exact way. In all normed spaces, the topology \textbf{GC: , and therefore the $\sigma$-algebra,} is constructed from sets of the form ``the object is within $\epsilon$ of $x$''.

@@

Proof. Let $p \in L(H)$ be a closed sub-space of $H$. Then $p$ is a closed set, the complement of an open set, and therefore a Borel set: $p \in \Sigma$. Let $v \in H$ be a vector in the Hilbert space and consider the singleton $U = \{ v \}$. This is not a closed sub-space and therefore $U \notin L(H)$. As the topology of a Hilbert space is Hausdorff, all singletons are closed and therefore $U \in \Sigma$.\footnote{To give more context, the set $U = \{ v \}$ can be written as $U = \{w \; | \; |w - v|^2 = 0\}$ the set of all elements with zero distance from $v$. This can be understood as the limit (i.e. countable intersection) of a sequence of propositions ``the object is within $\epsilon$ of $v$'' where $\epsilon$ tends to zero. \textbf{GC: Ho messo come note, cos\`i la dimostrazione rimane visibilmente piccola. Vedi te se metterla nel testo.} } This means $L(H) \subset \Sigma$. This concludes the proof.

@@

Now that we have briefly seen the definitions and basic constructions, we can gain more insights into these structures and what they represent. First, let us understand why the lattice of subspaces (i.e. quantum logic) fails to be a Boolean algebra. Stone's representation theorem \textbf{GC: serve citazione qui? Si pu\`o fare Davey... Oppure il paper originale... } tells us that any Boolean algebra can be expressed as an algebra of sets using the standard set operations (i.e. intersection, union and complement). For the lattice of subspaces, note that the disjunction $p \vee q$ does not correspond to the set union of the subspaces, but the span. In fact, the union of subspaces is not part of the lattice.  This is why the lattice, even if it is a complemented $0-1$ lattice, fails to be a Boolean algebra: it does not use the standard set operations.

@@

We want to be clear that we are not claiming that the lattice of subspaces of quantum logic is not of interest, or that it is of equal interest in both classical and quantum mechanics. Here we have pointed \textbf{GC: tolto "it" } out that the argument that quantum mechanics \emph{requires} necessarily a radical departure from classical logic is untenable, given that both theories are built on similar structures. We conclude that such claims should be substantially weakened, in the light of the arguments that we proposed in this section.
